Variant calling with transformers

### 2022-05-23 manuscript/manuscript.tex
Try some BAM feats for classifier

All variant discovery tools must address two challenges. First, they must identify candidate alleles in a region. Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine the most likely alleles. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles. Later tools (e.g. Poplin et al. 2018, Kim et al. 2018, Cooke et al. 2021) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).

@@

Maybe brief explanation of how transformers work?

Here, we explore a new approach to variant detection that replaces the bespoke statistical techniques used in previous tools with a single, end-to-end machine learning model. We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.

@@

>>>>

All variant discovery tools must address two challenges. First, callers must identify potential  alleles in a region. Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine which should be included in the caller output. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles (Nielsen et al. 2011). Later tools (e.g. Poplin et al. 2018, Kim et al. 2018, Cooke et al. 2021) incorporated local re-assembly of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).

@@

While replacing the statistical and heuristic components of the allele likelihood calculation with a deep learning model has resulted in improved specificity of variant callers, current tools continue to rely on complex, handcrafted statistical machinery to identify possible alleles. Here, we describe a new approach that replaces both the candidate allele generation and the likelihood calculation with a single, learned model. We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.

@@

\vspace{8pt}
Nielsen, Rasmus, et al. "Genotype and SNP calling from next-generation sequencing data." Nature Reviews Genetics 12.6 (2011): 443-451.
